A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-09 tidyindex.tex
Update section 2

The tidy framework consists of two key components: tidy data and tidy
tools. The concept of tidy data (Wickham 2014) prescribes specific rules
for organising data in an analysis, with observations as rows, variables
as columns, and types of observational units as tables. Tidy tools, on
the other hand, are concatenated in a sequence through which the tidy
object flows, creating a pipeline for data processing. These pipelines
are data centric, meaning all the functions take a tidy data object as
inputs and return a processed tidy data object, directly ready for the
next operations to be applied. Also, the pipeline approach corresponds
to the modular programming practice, which breaks down complex problems
into smaller and more manageable pieces, as oppose to a monolithic
design, where all the steps are predetermined and integrated into a
single piece. The flexibility provided by the modularity makes it easier
to modify certain steps in the pipeline and to maintain and extend the
code base.

Examples of using a pipeline approach to structure data analysis in R
can be traced back to the interactive graphics literature in the earlier
time (Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng
2014; Wickham et al. 2009). Wickham et al. (2009) argues that whether
made explicit or not, pipeline has to be presented in every graphics
program and making them explicit is beneficial for understanding the
implementation and comparing between different graphic systems. While
this comment is made in the context of graphics program, it is also
applicable generally to data analysis workflow. More recently, the
tidyverse suite (Wickham et al. 2019) takes the pipeline approach for
general-purpose data wrangling and has gained popularity within the R
community. The pipeline-style code can be directly read as a series of
operations applied successively on the tidy data object, offering a
method to document the data wrangling process for analysts to
comprehend. Unlike textual documentation, which may omit seemingly
obvious or less significant information, the pipeline approach captures
all the computational details while retaining conciseness.

Since the success of tidyverse, more packages have been developed to
analyze data using the tidy framework for domains specific applications,
a noticeable example of which is \texttt{tidymodels} for building
machine learning models (Kuhn and Wickham 2020). To create a tidy
workflow tailored to a specific domain, developers first need to
identify the fundamental building blocks in the typical workflow. These
components are then implemented as modules, which can be combined to
form the pipeline. For example, in supervised machine learning models,
steps such as data splitting, model training and model evaluation are
commonly used in most workflow. In the \texttt{tidymodels}, these steps
are correspondingly implemented as package \texttt{rsample},
\texttt{parsnip}, and \texttt{yardstick}, agnostic to the specific model
chosen. The uniform interface in tidymodels frees analysts from
recalling model-specific syntax for performing the same operation across
different models, increasing the efficiency to work with different
models simultaneously.

For constructing indexes, rather than implementing each index
separately, as has been done for some indexes, the pipeline approach
introduces explicit, user-facing, and standalone modules for analysts to
assemble indexes. Analysts can rearrange the order in the pipeline or
insert new step to experiment with possible new index definition or to
understand index behaviour. Also, the readability of the pipeline code
can complement conventional diagrams and textual information to
understand the how indexes are constructed. The uniform interface of
pipeline allows analysts to assemble a diverse array of indexes with a
shared set of components, reducing the learning curve for adopting new
implementation for different indexes.

>>>>

The tidy framework consists of two key components: tidy data and tidy
tools. The concept of tidy data (Wickham 2014) prescribes specific rules
for organising data in an analysis, with observations as rows, variables
as columns, and types of observational units as tables. Tidy tools, on
the other hand, are concatenated in a sequence through which the tidy
data flows, creating a pipeline for data processing and modelling. These pipelines
are data centric, meaning all the tidy tools or functions take a tidy data object as
input and return a processed tidy data object, directly ready for the
next operations to be applied. Also, the pipeline approach corresponds
to the modular programming practice, which breaks down complex problems
into smaller and more manageable pieces.
% , as oppose to a monolithic
% design, where all the steps are predetermined and integrated into a
% single piece. 
The flexibility provided by the modularity makes it easier
to modify any step in the pipeline and to maintain and extend the
code base.

Examples of using a pipeline approach for data analysis 
can be traced back to the interactive graphics literature including 
Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng
2014; Wickham et al. 2009). Wickham et al. (2009) argues that whether
made explicit or not, a pipeline has to be presented in every graphics
program and making them explicit is beneficial for understanding the
implementation and comparing between different graphic systems. While
this comment is made in the context of graphics program (WHAT DO YOU MEAN by graphics program?), it is also
applicable generally to any data analysis workflow. More recently, the
tidyverse suite (Wickham et al. 2019) takes the pipeline approach for
general-purpose data wrangling and has gained popularity within the R
community. The pipeline-style code can be directly read as a series of
operations applied successively on a tidy data object, offering a
method to create a reproducible data wrangling process that is also easy to document.
% Unlike textual documentation, which may omit seemingly
% obvious or less significant information, the pipeline approach captures
% all the computational details while retaining conciseness.

Since the success of tidyverse, more packages have been developed to
analyze data using the tidy framework for domains specific applications,
a noticeable example of which is \texttt{tidymodels} for building
machine learning models (Kuhn and Wickham 2020). To create a tidy
workflow tailored to a specific domain, developers first need to
identify the fundamental building blocks to create a workflow. These
components are then implemented as modules, which can be combined to
form the pipeline. For example, in supervised machine learning models,
steps such as data splitting, model training and model evaluation are
commonly used in most workflow. In the \texttt{tidymodels}, these steps
are correspondingly implemented as package \texttt{rsample},
\texttt{parsnip}, and \texttt{yardstick}, agnostic to the specific model
chosen. The uniform interface in tidymodels frees analysts from
recalling model-specific syntax for performing the same operation across
different models, increasing the efficiency to work with different
models simultaneously.

For constructing indexes, rather than implementing each index
separately, as it has been done for some indexes, the pipeline approach
introduces explicit, user-facing (??), and standalone modules for analysts to
assemble indexes. Analysts can then rearrange the pipeline or
insert new steps to create indexes and/or to experiment with possible index definition while also testing
the index behaviour. 
% 
% In addition, the readability of the pipeline code
% can complement conventional diagrams and textual information to
% understand the how indexes are constructed.

(I don't think this is needed -->)The uniform interface of a
pipeline allows analysts to assemble a diverse array of indexes with a
shared set of components, reducing the learning curve for adopting new
implementation for different indexes.
